---
title: "AML Multi-omic Analysis - Section 2 (TE Volcano: Discovery + Validation)"
author: "Research Team"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  dev = "png",
  fig.retina = 2
)

suppressPackageStartupMessages({
  library(DESeq2)
  library(dplyr)
  library(tibble)
  library(ggplot2)
  library(gridExtra)
  library(ggrepel)
})
```

## 1) Auto-download input files (edit URLs)

Replace the placeholder URLs with your real file links (HTTP/HTTPS).  
If you already have files locally, set `download_inputs <- FALSE`.

```{r download-inputs}
download_inputs <- TRUE

# ---- Put your real URLs here ----
url_gene_c1 <- "BEAT_BM_c1_gene_counts.txt"
url_gene_c2 <- "BEAT_BM_c2_gene_counts.txt"
url_te_rda  <- "BEAT_BM_TE_subfamily_counts.Rda"
url_meta_c1 <- "aml_baseline_clinical_beats_BM_c1_cyto.csv"
url_meta_c2 <- "aml_baseline_clinical_beats_BM_c2_cyto.csv"

# ---- Local filenames (saved into ./data) ----
dir.create("data", showWarnings = FALSE, recursive = TRUE)

path_gene_c1 <- "BEAT_BM_c1_gene_counts.txt"
path_gene_c2 <- "BEAT_BM_c2_gene_counts.txt"
path_te_rda  <- "BEAT_BM_TE_subfamily_counts.Rda"
path_meta_c1 <- "aml_baseline_clinical_beats_BM_c1_cyto.csv"
path_meta_c2 <- "aml_baseline_clinical_beats_BM_c2_cyto.csv"

dl <- function(url, dest) {
  if (is.na(url) || url == "" || grepl("YOUR_LINK", url)) {
    stop("Please replace placeholder URLs (YOUR_LINK) with real links before running.")
  }
  if (!file.exists(dest)) {
    message("Downloading: ", basename(dest))
    utils::download.file(url, destfile = dest, mode = "wb", quiet = TRUE)
  } else {
    message("Found local file: ", dest)
  }
}

if (isTRUE(download_inputs)) {
  dl(url_gene_c1, path_gene_c1)
  dl(url_gene_c2, path_gene_c2)
  dl(url_te_rda,  path_te_rda)
  dl(url_meta_c1, path_meta_c1)
  dl(url_meta_c2, path_meta_c2)
}

stopifnot(file.exists(path_gene_c1), file.exists(path_gene_c2),
          file.exists(path_te_rda),  file.exists(path_meta_c1),
          file.exists(path_meta_c2))
```

## 2) Configuration

```{r config}
target_cols <- c(
  "dbgap_rnaseq_sample", "response_induction", "sex", "age_dx",
  "npm1", "runx1", "asxl1", "tp53", "DNMT3A", "IDH1", "IDH2",
  "KRAS", "NRAS", "PTPN11", "spliceosome", "cohesin", "ras_pathway",
  "cyto_t_8_21", "cyto_inv16", "cyto_cbf", "cyto_t_15_17",
  "cyto_inv3", "cyto_t_6_9", "cyto_5_7_abn", "cyto_plus8",
  "cyto_normal", "eln_clean"
)
```

## 3) Load TE counts from `.Rda`

```{r load-te}
te_env <- new.env()
load(path_te_rda, envir = te_env)

te_obj_name <- ls(te_env)[sapply(ls(te_env), function(x)
  is.matrix(get(x, te_env)) || is.data.frame(get(x, te_env))
)][1]

if (is.na(te_obj_name) || length(te_obj_name) == 0) {
  stop("No matrix/data.frame object found inside the TE .Rda file.")
}

te_df <- as.data.frame(get(te_obj_name, envir = te_env), check.names = FALSE)

if (!("TE_subfamily" %in% colnames(te_df))) {
  stop("Expected a column named 'TE_subfamily' in TE object.")
}

te_mat <- as.matrix(as.data.frame(lapply(
  te_df[, colnames(te_df) != "TE_subfamily", drop = FALSE],
  function(x) as.numeric(trimws(as.character(x)))
)))

rownames(te_mat) <- paste0("TEF_", te_df$TE_subfamily)

message("TE matrix dims: ", nrow(te_mat), " x ", ncol(te_mat))
```

## 4) Helper functions

```{r helpers}
prep_meta <- function(path_meta, target_cols) {
  read.csv(path_meta) %>%
    filter(response_induction %in% c("Complete Response", "Refractory")) %>%
    select(all_of(target_cols)) %>%
    mutate(
      response_induction = factor(gsub(" ", "_", response_induction),
                                  levels = c("Complete_Response", "Refractory")),
      sex = factor(sex)
    ) %>%
    column_to_rownames("dbgap_rnaseq_sample")
}

run_deseq <- function(gene_counts_path, meta_df, te_mat) {
  gene <- read.delim(gene_counts_path, header = TRUE, row.names = 1, check.names = FALSE)

  common <- base::intersect(base::intersect(colnames(gene), colnames(te_mat)), rownames(meta_df))
  if (length(common) < 5) stop("Too few overlapping samples after intersect(). Check sample IDs.")

  counts <- rbind(gene[, common, drop = FALSE], te_mat[, common, drop = FALSE])

  dds <- DESeqDataSetFromMatrix(
    countData = round(counts),
    colData   = meta_df[common, , drop = FALSE],
    design    = ~ sex + age_dx + response_induction
  )

  dds <- DESeq(dds, quiet = TRUE)
  res <- results(dds, contrast = c("response_induction", "Refractory", "Complete_Response"))
  list(dds = dds, res = res, common = common)
}

prep_te_res <- function(res) {
  as.data.frame(res) %>%
    rownames_to_column("feature") %>%
    filter(grepl("TEF_LTR|TEF_LINE|TEF_SINE", feature)) %>%
    mutate(
      te_class = case_when(
        grepl("LTR", feature)  ~ "LTR",
        grepl("LINE", feature) ~ "LINE",
        grepl("SINE", feature) ~ "SINE",
        TRUE ~ "Other"
      ),
      feature = gsub("TEF_", "", feature),
      status = case_when(
        !is.na(pvalue) & pvalue < 0.01 & log2FoldChange > 0 ~ "Upregulated",
        !is.na(pvalue) & pvalue < 0.01 & log2FoldChange < 0 ~ "Downregulated",
        TRUE ~ "NS"
      ),
      neglog10p = ifelse(is.na(pvalue), NA_real_, -log10(pvalue))
    )
}

volcano_plot <- function(df,
                         title,
                         lfc_thr_label = 0.5,
                         p_thr_label   = 0.01,
                         max_labels    = 30,
                         drop_class_prefix = TRUE,
                         y_lim = c(-0.1, 6)) {

  stopifnot(all(c("pvalue","log2FoldChange","neglog10p","status","feature") %in% colnames(df)))

  plot_df <- df %>%
    dplyr::mutate(
      feature_clean = if (drop_class_prefix) sub("^[^:]+:", "", feature) else feature,
      label_me = !is.na(pvalue) & pvalue < p_thr_label & log2FoldChange > lfc_thr_label
    )

  label_df <- plot_df %>%
    dplyr::filter(label_me) %>%
    dplyr::arrange(dplyr::desc(neglog10p)) %>%
    dplyr::slice_head(n = max_labels)

  ggplot2::ggplot(plot_df, ggplot2::aes(x = log2FoldChange, y = neglog10p)) +
    ggplot2::geom_point(ggplot2::aes(color = status),
                        alpha = 0.75, size = 1.8, na.rm = TRUE) +
    ggrepel::geom_text_repel(
      data = label_df,
      ggplot2::aes(label = feature_clean),
      size = 3.0,
      fontface = "bold",
      max.overlaps = Inf,
      box.padding = 0.30,
      point.padding = 0.20,
      min.segment.length = 0,
      segment.size = 0.25
    ) +
    ggplot2::geom_vline(
      xintercept = c(-lfc_thr_label, lfc_thr_label),
      linetype = "dashed", linewidth = 0.35
    ) +
    ggplot2::geom_hline(
      yintercept = -log10(p_thr_label),
      linetype = "dashed", linewidth = 0.35
    ) +
    ggplot2::scale_color_manual(
      values = c("Upregulated" = "#D55E00",
                 "Downregulated" = "#0072B2",
                 "NS" = "grey70"),
      name = NULL
    ) +
    ggplot2::labs(
      title = title,
      subtitle = paste0(
        "Labels: upregulated (log2FC>", lfc_thr_label,
        ", p<", p_thr_label, "); top ", max_labels, " by -log10(p)"
      ),
      x = "log2(Fold Change)",
      y = "-log10(p-value)"
    ) +
    ggplot2::theme_classic(base_size = 12) +
    ggplot2::theme(
      legend.position = "top",
      plot.title = ggplot2::element_text(face = "bold", size = 14),
      plot.subtitle = ggplot2::element_text(face = "bold", size = 10),
      axis.title = ggplot2::element_text(face = "bold", size = 12),
      axis.text  = ggplot2::element_text(face = "bold", size = 10),
      axis.line  = ggplot2::element_line(linewidth = 0.6),
      axis.ticks = ggplot2::element_line(linewidth = 0.6),
      panel.grid = ggplot2::element_blank()
    ) +
    ggplot2::coord_cartesian(ylim = y_lim)
}

```

## 5) Discovery cohort (c1)

```{r discovery}
meta_c1 <- prep_meta(path_meta_c1, target_cols)
fit_c1  <- run_deseq(path_gene_c1, meta_c1, te_mat)
res_te_c1 <- prep_te_res(fit_c1$res)

if (nrow(res_te_c1) == 0) stop("No TE rows found in Discovery cohort after filtering!")

p_c1 <- volcano_plot(res_te_c1, "Discovery (Wave 1+2)")
print(p_c1)
```

## 6) Validation cohort (c2)

```{r validation}
meta_c2 <- prep_meta(path_meta_c2, target_cols)
fit_c2  <- run_deseq(path_gene_c2, meta_c2, te_mat)
res_te_c2 <- prep_te_res(fit_c2$res)

p_c2 <- volcano_plot(res_te_c2, "Validation (Wave 3+4)")
print(p_c2)
```

## 7) Side-by-side comparison

```{r combined, fig.width=12, fig.height=5}
gridExtra::grid.arrange(p_c1, p_c2, ncol = 2)
```

## 8) Save outputs (tables + plots)

```{r save-outputs}
dir.create("outputs", showWarnings = FALSE, recursive = TRUE)

write.csv(res_te_c1, file.path("outputs", "TE_results_discovery_c1.csv"), row.names = FALSE)
write.csv(res_te_c2, file.path("outputs", "TE_results_validation_c2.csv"), row.names = FALSE)

ggsave(file.path("outputs", "volcano_discovery_c1.png"), plot = p_c1, width = 8, height = 5, dpi = 200)
ggsave(file.path("outputs", "volcano_validation_c2.png"), plot = p_c2, width = 8, height = 5, dpi = 200)

message("Saved outputs to: ./outputs")
```


```{r pies-and-venn-upregulated-families, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(forcats)
  library(VennDiagram)
  library(grid)
})

# -----------------------------
# Settings
# -----------------------------
p_thr   <- 0.01
lfc_thr <- 0

families_keep <- c(
  "LTR:ERV1","LTR:ERVL","LTR:ERVL-MaLR","LTR:ERVK",
  "SINE:RTE-X","LINE:L1","LINE:CR1","SINE:Alu","SINE:MIR"
)

# -----------------------------
# Helper: parse feature -> Family_ID + Subfamily key
# Assumes feature looks like "LTR:ERVK:XYZ" (or similar).
# If your feature format is different, tell me and I’ll adapt the parser.
# -----------------------------
prep_up <- function(res_te, cohort_name) {
  res_te %>%
    mutate(
      Cohort = cohort_name,
      TE = feature
    ) %>%
    separate(TE, into = c("Class","Family","Subfamily"), sep=":",
             remove = FALSE, fill = "right", extra = "merge") %>%
    mutate(
      Class = coalesce(Class, "Other"),
      Family = coalesce(Family, "Other"),
      Subfamily = coalesce(Subfamily, TE),
      Family_ID = paste(Class, Family, sep=":"),
      # unique key for overlap testing
      SubKey = paste(Class, Family, Subfamily, sep=":")
    ) %>%
    filter(!is.na(pvalue), !is.na(log2FoldChange)) %>%
    filter(pvalue < p_thr, log2FoldChange > lfc_thr) %>%
    filter(Family_ID %in% families_keep)
}

up_c1 <- prep_up(res_te_c1, "Discovery (c1)")
up_c2 <- prep_up(res_te_c2, "Validation (c2)")

# -----------------------------
# PIE CHARTS: Contribution by Family_ID
# -----------------------------
pie_df <- bind_rows(up_c1, up_c2) %>%
  count(Cohort, Family_ID, name = "n_up") %>%
  group_by(Cohort) %>%
  mutate(
    pct = 100 * n_up / sum(n_up),
    Family_ID = factor(Family_ID, levels = families_keep)
  ) %>%
  ungroup()

pie_plot <- function(cohort_name) {
  d <- pie_df %>% filter(Cohort == cohort_name) %>% arrange(desc(pct))
  if (nrow(d) == 0) {
    return(ggplot() + theme_void() +
             labs(title = paste0("Upregulated contribution: ", cohort_name),
                  subtitle = "No upregulated subfamilies passed filters in selected families"))
  }

  ggplot(d, aes(x = "", y = n_up, fill = Family_ID)) +
    geom_col(width = 1, colour = "white") +
    coord_polar(theta = "y") +
    theme_void() +
    labs(
      title = paste0("Upregulated TE subfamilies by family: ", cohort_name),
      subtitle = paste0("p < ", p_thr, ", log2FC > ", lfc_thr),
      fill = "Family"
    ) +
    geom_text(
      aes(label = sprintf("%.1f%%", pct)),
      position = position_stack(vjust = 0.5),
      size = 3,
      fontface = "bold"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(face = "bold", size = 10),
      legend.title = element_text(face = "bold"),
      legend.text = element_text(face = "bold")
    )
}

p_pie_c1 <- pie_plot("Discovery (c1)")
p_pie_c2 <- pie_plot("Validation (c2)")
# 
# print(p_pie_c1)
# print(p_pie_c2)

# Save pies
dir.create("outputs", showWarnings = FALSE, recursive = TRUE)
ggsave("outputs/Pie_upregulated_family_contribution_Discovery.png", p_pie_c1, width = 7, height = 6, dpi = 200)
ggsave("outputs/Pie_upregulated_family_contribution_Validation.png", p_pie_c2, width = 7, height = 6, dpi = 200)

# -----------------------------
# VENN: overlap of upregulated subfamilies (SubKey)
# -----------------------------
set_c1 <- unique(up_c1$SubKey)
set_c2 <- unique(up_c2$SubKey)

venn_obj <- VennDiagram::venn.diagram(
  x = list(`Discovery (c1)` = set_c1, `Validation (c2)` = set_c2),
  filename = NULL,   # draw to R device
  fill = c("#56B4E9", "#E69F00"),
  alpha = 0.5,
  cex = 1.3,
  cat.cex = 1.1,
  cat.fontface = "bold",
  fontface = "bold",
  margin = 0.1
)

grid::grid.newpage()
grid::grid.draw(venn_obj)

# Save Venn as PNG
png("outputs/Venn_upregulated_subfamilies_Discovery_vs_Validation.png", width = 900, height = 700, res = 150)
grid::grid.newpage()
grid::grid.draw(venn_obj)
dev.off()

# Also print counts to console/table
overlap_n <- length(intersect(set_c1, set_c2))
only_c1   <- length(setdiff(set_c1, set_c2))
only_c2   <- length(setdiff(set_c2, set_c1))

cat("\nUpregulated overlap counts (within selected families):\n")
cat("Discovery only:", only_c1, "\n")
cat("Validation only:", only_c2, "\n")
cat("Shared:", overlap_n, "\n")
```






```{r shared-upregulated-lfc-and-vst-boxplots, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(DESeq2)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(forcats)
})

# -----------------------------
# Settings
# -----------------------------
p_thr   <- 0.01
lfc_thr <- 0
max_shared_to_plot <- 9

families_keep <- c(
  "LTR:ERV1","LTR:ERVL","LTR:ERVL-MaLR","LTR:ERVK",
  "SINE:RTE-X","LINE:L1","LINE:CR1","SINE:Alu","SINE:MIR"
)

# -----------------------------
# Helper: upregulated TE set with stable key
# (No need for dds objects here)
# -----------------------------
prep_up_set <- function(res_te, cohort_name) {
  res_te %>%
    dplyr::mutate(Cohort = cohort_name, TE = feature) %>%
    tidyr::separate(
      TE, into = c("Class","Family","Subfamily"),
      sep=":", remove = FALSE, fill="right", extra="merge"
    ) %>%
    dplyr::mutate(
      Class     = dplyr::coalesce(Class, "Other"),
      Family    = dplyr::coalesce(Family, "Other"),
      Subfamily = dplyr::coalesce(Subfamily, TE),
      Family_ID = paste(Class, Family, sep=":"),
      SubKey    = paste(Class, Family, Subfamily, sep=":")
    ) %>%
    dplyr::filter(
      !is.na(pvalue), !is.na(log2FoldChange),
      pvalue < p_thr, log2FoldChange > lfc_thr,
      Family_ID %in% families_keep
    )
}

up_c1 <- prep_up_set(res_te_c1, "Discovery (c1)")
up_c2 <- prep_up_set(res_te_c2, "Validation (c2)")

shared_keys <- intersect(unique(up_c1$SubKey), unique(up_c2$SubKey))

if (length(shared_keys) == 0) {
  cat("No shared upregulated TE subfamilies between Discovery and Validation at p<",
      p_thr, " and log2FC>", lfc_thr, " within selected families.\n", sep="")
} else {

  # -----------------------------
  # 1) Bar chart: log2FC Discovery vs Validation for shared
  # -----------------------------
  lfc_c1 <- up_c1 %>%
    dplyr::filter(SubKey %in% shared_keys) %>%
    dplyr::select(SubKey, Class, Family, Subfamily, log2FoldChange) %>%
    dplyr::rename(log2FC_Discovery = log2FoldChange)

  lfc_c2 <- up_c2 %>%
    dplyr::filter(SubKey %in% shared_keys) %>%
    dplyr::select(SubKey, log2FoldChange) %>%
    dplyr::rename(log2FC_Validation = log2FoldChange)

  shared_lfc <- lfc_c1 %>%
    dplyr::inner_join(lfc_c2, by = "SubKey") %>%
    dplyr::mutate(mean_log2FC = (log2FC_Discovery + log2FC_Validation) / 2) %>%
    dplyr::arrange(dplyr::desc(mean_log2FC)) %>%
    dplyr::slice_head(n = max_shared_to_plot)

  shared_lfc_long <- shared_lfc %>%
    tidyr::pivot_longer(
      cols = c(log2FC_Discovery, log2FC_Validation),
      names_to = "Cohort",
      values_to = "log2FC"
    ) %>%
    dplyr::mutate(
      Cohort = dplyr::recode(Cohort,
                             log2FC_Discovery = "Discovery (c1)",
                             log2FC_Validation = "Validation (c2)"),
      SubLabel = paste0(Class, ":", Family, ":", Subfamily),
      SubLabel = factor(SubLabel, levels = unique(SubLabel))
    )

  p_lfc <- ggplot(shared_lfc_long, aes(x = SubLabel, y = log2FC, fill = Cohort)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    labs(
      title = "Shared upregulated TE subfamilies: log2FC (Discovery vs Validation)",
      subtitle = paste0("p<", p_thr, ", log2FC>", lfc_thr,
                        " | top ", min(max_shared_to_plot, nrow(shared_lfc)), " by mean log2FC"),
      x = "TE subfamily",
      y = "log2(Fold Change)"
    ) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 60, hjust = 1, face = "bold", size = 9),
      axis.text.y = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(face = "bold"),
      legend.position = "top",
      legend.title = element_blank()
    )

  # print(p_lfc)

  dir.create("outputs", showWarnings = FALSE, recursive = TRUE)
  ggsave("outputs/Shared_upregulated_log2FC_Discovery_vs_Validation.png",
         p_lfc, width = 13, height = 6, dpi = 200)

  # -----------------------------
  # 2) Boxplots: VST expression by response (FAST)
  # Requires dds_c1 / dds_c2. If not present, we skip boxplots gracefully.
  # -----------------------------
  if (!exists("dds_c1") || !exists("dds_c2")) {
    cat("dds_c1/dds_c2 not found in this Rmd run, so skipping VST boxplots.\\n",
        "Run the DESeq2 sections that create dds_c1 and dds_c2 before this chunk.\\n", sep="")
  } else {

    shared_top_keys <- shared_lfc$SubKey
    te_rows <- paste0("TEF_", shared_top_keys)

    vsd_c1 <- DESeq2::vst(dds_c1, blind = TRUE)
    vsd_c2 <- DESeq2::vst(dds_c2, blind = TRUE)

    get_box_df <- function(vsd, cohort_name, te_rows_keep) {
      mat <- SummarizedExperiment::assay(vsd)
      present <- intersect(rownames(mat), te_rows_keep)
      if (length(present) == 0) return(NULL)

      expr <- mat[present, , drop = FALSE] %>%
        as.data.frame() %>%
        tibble::rownames_to_column("Feature")

      long <- expr %>%
        tidyr::pivot_longer(-Feature, names_to = "Sample", values_to = "VST") %>%
        dplyr::mutate(SubKey = sub("^TEF_", "", Feature), Cohort = cohort_name)

      meta <- as.data.frame(SummarizedExperiment::colData(vsd)) %>%
        tibble::rownames_to_column("Sample")

      long %>%
        dplyr::left_join(meta, by = "Sample") %>%
        dplyr::filter(!is.na(response_induction)) %>%
        dplyr::mutate(
          response_induction = factor(response_induction,
                                       levels = c("Complete_Response", "Refractory")),
          TE_label = factor(SubKey, levels = shared_top_keys)
        )
    }

    box_df <- dplyr::bind_rows(
      get_box_df(vsd_c1, "Discovery (c1)", te_rows),
      get_box_df(vsd_c2, "Validation (c2)", te_rows)
    )

    if (is.null(box_df) || nrow(box_df) == 0) {
      cat("Could not extract VST expression for shared TE rows from dds objects.\\n")
    } else {

      p_box <- ggplot(box_df, aes(x = response_induction, y = VST, fill = response_induction)) +
        geom_boxplot(outlier.size = 0.6, width = 0.7) +
        facet_grid(TE_label ~ Cohort, scales = "free_y") +
        labs(
          title = "Shared upregulated TE subfamilies: expression by response",
          subtitle = "VST expression | Complete Response vs Refractory",
          x = NULL,
          y = "VST expression"
        ) +
        theme_classic(base_size = 12) +
        theme(
          strip.text = element_text(face = "bold", size = 9),
          axis.text.x = element_text(face = "bold", angle = 20, hjust = 1),
          axis.text.y = element_text(face = "bold"),
          axis.title = element_text(face = "bold"),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(face = "bold"),
          legend.position = "none"
        )

      # print(p_box)

      ggsave(
        "outputs/Shared_upregulated_VST_boxplots_Response_byCohort.png",
        p_box,
        width = 12,
        height = max(6, 0.35 * length(unique(box_df$TE_label)) + 3),
        dpi = 200
      )
    }
  }
}


```


```{r shared-upregulated-vst-boxplots-force-new, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(DESeq2)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(forcats)
})

# -----------------------------
# Settings
# -----------------------------
p_thr   <- 0.01
lfc_thr <- 0
max_shared_to_plot <- 9

families_keep <- c(
  "LTR:ERV1","LTR:ERVL","LTR:ERVL-MaLR","LTR:ERVK",
  "SINE:RTE-X","LINE:L1","LINE:CR1","SINE:Alu","SINE:MIR"
)

# -----------------------------
# Shared upregulated TE subfamilies (keys)
# -----------------------------
prep_up_set <- function(res_te, cohort_name) {
  res_te %>%
    mutate(Cohort = cohort_name, TE = feature) %>%
    separate(TE, into = c("Class","Family","Subfamily"), sep=":",
             remove = FALSE, fill="right", extra="merge") %>%
    mutate(
      Class     = coalesce(Class, "Other"),
      Family    = coalesce(Family, "Other"),
      Subfamily = coalesce(Subfamily, TE),
      Family_ID = paste(Class, Family, sep=":"),
      SubKey    = paste(Class, Family, Subfamily, sep=":")
    ) %>%
    filter(!is.na(pvalue), !is.na(log2FoldChange),
           pvalue < p_thr, log2FoldChange > lfc_thr,
           Family_ID %in% families_keep)
}

up_c1 <- prep_up_set(res_te_c1, "Discovery (c1)")
up_c2 <- prep_up_set(res_te_c2, "Validation (c2)")

shared_keys <- intersect(unique(up_c1$SubKey), unique(up_c2$SubKey))
if (length(shared_keys) == 0) {
  stop("No shared upregulated TE subfamilies found with current thresholds.")
}

# Use the top shared by mean log2FC for readability (same as before)
lfc_c1 <- up_c1 %>% filter(SubKey %in% shared_keys) %>%
  select(SubKey, log2FoldChange) %>% rename(fc1 = log2FoldChange)
lfc_c2 <- up_c2 %>% filter(SubKey %in% shared_keys) %>%
  select(SubKey, log2FoldChange) %>% rename(fc2 = log2FoldChange)

shared_top_keys <- lfc_c1 %>%
  inner_join(lfc_c2, by = "SubKey") %>%
  mutate(mean_fc = (fc1 + fc2)/2) %>%
  arrange(desc(mean_fc)) %>%
  slice_head(n = max_shared_to_plot) %>%
  pull(SubKey)

te_rows <- paste0("TEF_", shared_top_keys)

# -----------------------------
# Build counts + metadata if missing (so VST ALWAYS works)
# Requires: path_gene_c1/path_gene_c2/path_te_rda and the meta csv paths
# -----------------------------
if (!exists("te_mat")) {
  te_env <- new.env()
  load(path_te_rda, envir = te_env)
  te_obj_name <- ls(te_env)[sapply(ls(te_env), function(x)
    is.matrix(get(x, te_env)) || is.data.frame(get(x, te_env))
  )][1]
  te_df <- as.data.frame(get(te_obj_name, envir = te_env), check.names = FALSE)
  te_mat <- as.matrix(as.data.frame(lapply(
    te_df[, colnames(te_df) != "TE_subfamily", drop = FALSE],
    function(x) as.numeric(trimws(as.character(x)))
  )))
  rownames(te_mat) <- paste0("TEF_", te_df$TE_subfamily)
}

if (!exists("counts_c1") || !exists("meta_c1")) {
  gene_c1_tmp <- read.delim(path_gene_c1, header = TRUE, row.names = 1, check.names = FALSE)
  meta_c1 <- read.csv(path_meta_c1) %>%
    filter(response_induction %in% c("Complete Response", "Refractory")) %>%
    select(all_of(target_cols)) %>%
    mutate(response_induction = factor(gsub(" ", "_", response_induction),
                                       levels = c("Complete_Response", "Refractory")),
           sex = factor(sex)) %>%
    column_to_rownames("dbgap_rnaseq_sample")

  common_c1 <- intersect(intersect(colnames(gene_c1_tmp), colnames(te_mat)), rownames(meta_c1))
  counts_c1 <- rbind(gene_c1_tmp[, common_c1, drop = FALSE], te_mat[, common_c1, drop = FALSE])
  meta_c1 <- meta_c1[common_c1, , drop = FALSE]
}

if (!exists("counts_c2") || !exists("meta_c2")) {
  gene_c2_tmp <- read.delim(path_gene_c2, header = TRUE, row.names = 1, check.names = FALSE)
  meta_c2 <- read.csv(path_meta_c2) %>%
    filter(response_induction %in% c("Complete Response", "Refractory")) %>%
    select(all_of(target_cols)) %>%
    mutate(response_induction = factor(gsub(" ", "_", response_induction),
                                       levels = c("Complete_Response", "Refractory")),
           sex = factor(sex)) %>%
    column_to_rownames("dbgap_rnaseq_sample")

  common_c2 <- intersect(intersect(colnames(gene_c2_tmp), colnames(te_mat)), rownames(meta_c2))
  counts_c2 <- rbind(gene_c2_tmp[, common_c2, drop = FALSE], te_mat[, common_c2, drop = FALSE])
  meta_c2 <- meta_c2[common_c2, , drop = FALSE]
}

# -----------------------------
# VST and boxplots (TEs only)
# -----------------------------
dds1 <- DESeqDataSetFromMatrix(round(counts_c1), meta_c1, design = ~ response_induction)
dds2 <- DESeqDataSetFromMatrix(round(counts_c2), meta_c2, design = ~ response_induction)

vsd1 <- varianceStabilizingTransformation(dds1, blind = TRUE)
vsd2 <- varianceStabilizingTransformation(dds2, blind = TRUE)

to_long <- function(vsd, cohort_name) {
  mat <- SummarizedExperiment::assay(vsd)
  present <- intersect(rownames(mat), te_rows)
  if (length(present) == 0) return(NULL)

  df <- mat[present, , drop = FALSE] %>%
    as.data.frame() %>%
    rownames_to_column("Feature") %>%
    pivot_longer(-Feature, names_to = "Sample", values_to = "VST") %>%
    mutate(
      SubKey = sub("^TEF_", "", Feature),
      Cohort = cohort_name
    )

  meta <- as.data.frame(SummarizedExperiment::colData(vsd)) %>%
    rownames_to_column("Sample")

  df %>%
    left_join(meta, by = "Sample") %>%
    mutate(
      response_induction = factor(response_induction,
                                  levels = c("Complete_Response", "Refractory")),
      TE_label = factor(SubKey, levels = shared_top_keys)
    )
}

box_df <- bind_rows(
  to_long(vsd1, "Discovery (c1)"),
  to_long(vsd2, "Validation (c2)")
)

if (is.null(box_df) || nrow(box_df) == 0) {
  stop("Boxplot data is empty: TE rows not found in VST matrices. Check TE feature naming (TEF_ prefix + Class:Family:Subfamily).")
}

p_box <- ggplot(box_df, aes(x = response_induction, y = VST, fill = response_induction)) +
  geom_boxplot(outlier.size = 0.6, width = 0.7) +
  facet_grid(TE_label ~ Cohort, scales = "free_y") +
  labs(
    title = "Shared upregulated TE subfamilies: expression by response",
    subtitle = "VST expression | Complete Response vs Refractory",
    x = NULL,
    y = "VST expression"
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_text(face = "bold", angle = 20, hjust = 1),
    axis.text.y = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(face = "bold"),
    legend.position = "none"
  )

# print(p_box)

dir.create("outputs", showWarnings = FALSE, recursive = TRUE)
ggsave("outputs/Shared_upregulated_VST_boxplots_Response_byCohort.png",
       p_box, width = 12,
       height = max(6, 0.35 * length(unique(box_df$TE_label)) + 3),
       dpi = 200)
```

```{r top_shared_upregulated_VST_jitter_pub, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(DESeq2)
  library(SummarizedExperiment)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(forcats)
  library(ggbeeswarm)
  library(rstatix)
})

# -----------------------------
# Settings
# -----------------------------
p_thr   <- 0.01
lfc_thr <- 0
top_n   <- 9

families_keep <- c(
  "LTR:ERV1","LTR:ERVL","LTR:ERVL-MaLR","LTR:ERVK",
  "SINE:RTE-X","LINE:L1","LINE:CR1","SINE:Alu","SINE:MIR"
)

# -----------------------------
# Helper: align counts and metadata (fixes DESeq2 ncol vs nrow errors)
# -----------------------------
align_counts_meta <- function(counts, meta, cohort = "Cohort") {
  counts <- as.matrix(counts)
  meta   <- as.data.frame(meta)

  # If meta has Sample column but rownames are not set, set them
  if (is.null(rownames(meta)) || any(rownames(meta) == "")) {
    if ("Sample" %in% colnames(meta)) rownames(meta) <- meta$Sample
    if ("sample" %in% colnames(meta)) rownames(meta) <- meta$sample
  }

  if (is.null(colnames(counts)) || any(colnames(counts) == "")) {
    stop(cohort, ": counts has missing/empty column names (sample IDs).")
  }
  if (is.null(rownames(meta)) || any(rownames(meta) == "")) {
    stop(cohort, ": meta has missing/empty rownames. Provide rownames(meta)=sample IDs or a Sample column.")
  }

  clean_id <- function(x) gsub("[[:space:]]+", "_", trimws(as.character(x)))
  colnames(counts) <- clean_id(colnames(counts))
  rownames(meta)   <- clean_id(rownames(meta))

  common <- intersect(colnames(counts), rownames(meta))
  if (length(common) == 0) {
    stop(cohort, ": no matching sample IDs between counts colnames and meta rownames.\n",
         "Example counts: ", paste(head(colnames(counts), 5), collapse=", "), "\n",
         "Example meta:   ", paste(head(rownames(meta), 5), collapse=", "))
  }

  if (length(common) < ncol(counts) || length(common) < nrow(meta)) {
    message(cohort, ": aligning on ", length(common), " shared samples (dropping non-shared).")
  }

  counts2 <- counts[, common, drop = FALSE]
  meta2   <- meta[common, , drop = FALSE]
  stopifnot(identical(colnames(counts2), rownames(meta2)))

  storage.mode(counts2) <- "integer"
  list(counts = counts2, meta = meta2)
}

# -----------------------------
# 0) Required objects check
# -----------------------------
need_objs <- c("res_te_c1","res_te_c2","counts_c1","meta_c1","counts_c2","meta_c2")
missing <- need_objs[!sapply(need_objs, exists)]
if (length(missing) > 0) {
  stop("Missing objects needed for this chunk: ", paste(missing, collapse = ", "),
       ".\nMake sure earlier chunks create these objects.")
}

# -----------------------------
# 1) Find TOP N shared upregulated subfamilies (Discovery ∩ Validation)
# -----------------------------
prep_up_set <- function(res_te, cohort_name) {
  res_te %>%
    mutate(Cohort = cohort_name, TE = feature) %>%
    separate(TE, into = c("Class","Family","Subfamily"), sep=":",
             remove = FALSE, fill="right", extra="merge") %>%
    mutate(
      Class     = coalesce(Class, "Other"),
      Family    = coalesce(Family, "Other"),
      Subfamily = coalesce(Subfamily, TE),
      Family_ID = paste(Class, Family, sep=":"),
      SubKey    = paste(Class, Family, Subfamily, sep=":")
    ) %>%
    filter(!is.na(pvalue), !is.na(log2FoldChange),
           pvalue < p_thr, log2FoldChange > lfc_thr,
           Family_ID %in% families_keep)
}

up_c1 <- prep_up_set(res_te_c1, "Discovery")
up_c2 <- prep_up_set(res_te_c2, "Validation")

shared_keys <- intersect(unique(up_c1$SubKey), unique(up_c2$SubKey))
if (length(shared_keys) == 0) stop("No shared upregulated subfamilies under current thresholds.")

top_keys <- up_c1 %>%
  filter(SubKey %in% shared_keys) %>%
  select(SubKey, log2FoldChange) %>%
  rename(fc1 = log2FoldChange) %>%
  inner_join(
    up_c2 %>%
      filter(SubKey %in% shared_keys) %>%
      select(SubKey, log2FoldChange) %>%
      rename(fc2 = log2FoldChange),
    by = "SubKey"
  ) %>%
  mutate(mean_fc = (fc1 + fc2) / 2) %>%
  arrange(desc(mean_fc)) %>%
  slice_head(n = top_n) %>%
  pull(SubKey)

te_rows <- paste0("TEF_", top_keys)

# -----------------------------
# 2) VST expression for TE rows (FAST, no DESeq run)
# -----------------------------
a1 <- align_counts_meta(counts_c1, meta_c1, cohort = "Discovery")
a2 <- align_counts_meta(counts_c2, meta_c2, cohort = "Validation")

dds1 <- DESeqDataSetFromMatrix(a1$counts, a1$meta, design = ~ response_induction)
dds2 <- DESeqDataSetFromMatrix(a2$counts, a2$meta, design = ~ response_induction)

vsd1 <- varianceStabilizingTransformation(dds1, blind = TRUE)
vsd2 <- varianceStabilizingTransformation(dds2, blind = TRUE)

to_long <- function(vsd, cohort_name) {
  mat <- SummarizedExperiment::assay(vsd)
  present <- intersect(rownames(mat), te_rows)
  if (length(present) == 0) {
    stop("None of the selected TE rows found in ", cohort_name,
         " VST matrix. Check TE rownames. Example rownames: ",
         paste(head(rownames(mat), 5), collapse = ", "))
  }

  df <- mat[present, , drop = FALSE] %>%
    as.data.frame() %>%
    rownames_to_column("Feature") %>%
    pivot_longer(-Feature, names_to = "Sample", values_to = "expr") %>%
    mutate(
      SubKey  = sub("^TEF_", "", Feature),
      Cohort  = cohort_name
    )

  meta <- as.data.frame(SummarizedExperiment::colData(vsd)) %>%
    rownames_to_column("Sample")

  df %>%
    left_join(meta, by = "Sample") %>%
    filter(response_induction %in% c("Complete_Response","Refractory")) %>%
    mutate(
      response_induction = factor(response_induction,
                                  levels = c("Complete_Response","Refractory"))
    )
}

plot_df <- bind_rows(
  to_long(vsd1, "Discovery"),
  to_long(vsd2, "Validation")
) %>%
  mutate(
    SubKey = factor(SubKey, levels = top_keys),
    Cohort = factor(Cohort, levels = c("Discovery","Validation"))
  )

# -----------------------------
# 3) Stats (Wilcoxon) + BH-FDR per cohort; y-position guaranteed
# -----------------------------
stats_df2 <- plot_df %>%
  group_by(Cohort, SubKey) %>%
  rstatix::wilcox_test(expr ~ response_induction) %>%
  ungroup() %>%
  group_by(Cohort) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  ungroup() %>%
  left_join(
    plot_df %>%
      group_by(Cohort, SubKey) %>%
      summarise(y = max(expr, na.rm = TRUE) + 0.15, .groups = "drop"),
    by = c("Cohort","SubKey")
  ) %>%
  mutate(
    p_lab = paste0("p=", signif(p, 2), "\nFDR=", signif(p.adj, 2)),
    x = "Refractory"
  )

# Strong median bars (explicit segments)
med_df <- plot_df %>%
  group_by(Cohort, SubKey, response_induction) %>%
  summarise(med = median(expr, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    x_num = as.numeric(factor(response_induction,
                              levels = levels(plot_df$response_induction))),
    x0 = x_num - 0.22,
    x1 = x_num + 0.22
  )

# -----------------------------
# 4) Publication-style jitter + median bars + p-values
# -----------------------------
# -----------------------------
# Clean publication jitter (properly centered median)
# -----------------------------

# ---- medians computed explicitly (robust) ----
med_df <- plot_df %>%
  group_by(Cohort, SubKey, response_induction) %>%
  summarise(med = median(expr, na.rm = TRUE), .groups = "drop")

dodge_w <- 0.65

p_jitter <- ggplot(
  plot_df,
  aes(x = SubKey, y = expr, color = response_induction)
) +
  # jitter / beeswarm (triangles, smaller)
  ggbeeswarm::geom_quasirandom(
    aes(group = response_induction),
    dodge.width = dodge_w,
    width = 0.12,
    alpha = 0.80,
    size = 0.85,
    shape = 17
  ) +

  # median “bar” as a horizontal tick (always visible + centered)
  geom_point(
    data = med_df,
    aes(x = SubKey, y = med, group = response_induction),
    inherit.aes = FALSE,
    position = position_dodge(width = dodge_w),
    shape = 95,          # horizontal line tick
    size = 7.5,          # length of the tick
    stroke = 0.15,       # thickness of the tick
    color = "black"
  ) +

  # p-values on top of each pair (one per SubKey x Cohort)
  geom_text(
    data = stats_df2,
    aes(x = SubKey, y = y, label = p_lab),
    inherit.aes = FALSE,
    size = 2.6,
    fontface = "bold",
    color = "black",
    vjust = 0
  ) +

  facet_grid(Cohort ~ ., scales = "free_y") +

  scale_color_manual(
    values = c("Complete_Response" = "#0072B2",
               "Refractory"        = "#D55E00"),
    name = "Response",
    labels = c("Complete response", "Refractory")
  ) +

  labs(
    title = "Top shared upregulated TE subfamilies",
    subtitle = "VST expression | triangles = samples | median in black | Wilcoxon p-value",
    x = "TE subfamily",
    y = "VST expression"
  ) +

  coord_cartesian(ylim = c(0, 15), clip = "off") +

  theme_classic(base_size = 10) +
  theme(
    strip.text = element_text(face = "bold", size = 8),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(face = "bold", size = 9),
    axis.title = element_text(face = "bold", size = 10),
    axis.text.x = element_text(face = "bold", size = 7, angle = 35, hjust = 1),
    axis.text.y = element_text(face = "bold", size = 7),
    axis.line  = element_line(linewidth = 0.6),
    axis.ticks = element_line(linewidth = 0.6),
    legend.title = element_text(face = "bold", size = 9),
    legend.text  = element_text(size = 8)
  )

# print(p_jitter)






# -----------------------------
# 4) Publication-style jitter + median ticks + significance stars (top of panel)
# -----------------------------

# -----------------------------
# Drop Class prefix (SINE/LINE/LTR) + medians + stars + plot
# -----------------------------

# 0) Clean SubKey labels for plotting (drop the first field before ":")
plot_df <- plot_df %>%
  mutate(
    SubKey_clean = sub("^[^:]+:", "", as.character(SubKey))
  )

# keep the original ordering (so panels stay in your intended top_keys order)
clean_levels <- sub("^[^:]+:", "", as.character(top_keys))
plot_df$SubKey_clean <- factor(plot_df$SubKey_clean, levels = clean_levels)

# 1) Medians per cohort x subfamily x response
med_df <- plot_df %>%
  group_by(Cohort, SubKey_clean, response_induction) %>%
  summarise(med = median(expr, na.rm = TRUE), .groups = "drop")

# 2) Significance stars per cohort x subfamily (uses stats_df2 with columns Cohort, SubKey, p)
stats_star <- stats_df2 %>%
  mutate(
    SubKey_clean = sub("^[^:]+:", "", as.character(SubKey)),
    stars = dplyr::case_when(
      p <= 0.0001 ~ "****",
      p <= 0.001  ~ "***",
      p <= 0.01   ~ "**",
      p <= 0.05   ~ "*",
      TRUE        ~ ""
    ),
    y = 14.2  # top of panel if using ylim 0–15
  )

# 3) Plot
dodge_w <- 0.65

p_jitter <- ggplot(
  plot_df,
  aes(x = SubKey_clean, y = expr, color = response_induction)
) +
  ggbeeswarm::geom_quasirandom(
    aes(group = response_induction),
    dodge.width = dodge_w,
    width = 0.12,
    alpha = 0.80,
    size = 0.75,
    shape = 17
  ) +
  # median tick
  geom_point(
    data = med_df,
    aes(x = SubKey_clean, y = med, group = response_induction),
    inherit.aes = FALSE,
    position = position_dodge(width = dodge_w),
    shape = 95,
    size = 6.0,     # tick length
    stroke = 0.35,  # tick thickness
    color = "black"
  ) +
  # stars at top
  geom_text(
    data = stats_star,
    aes(x = SubKey_clean, y = y, label = stars),
    inherit.aes = FALSE,
    size = 4.0,
    fontface = "bold",
    color = "black",
    vjust = 0
  ) +
  facet_grid(Cohort ~ ., scales = "free_y") +
  scale_color_manual(
    values = c("Complete_Response" = "#0072B2",
               "Refractory"        = "#D55E00"),
    name = "Response",
    labels = c("Complete response", "Refractory")
  ) +
  labs(
    title = "Top shared upregulated TE subfamilies",
    subtitle = "VST expression | triangles = samples | median in black | Wilcoxon significance",
    x = "TE subfamily",
    y = "VST expression"
  ) +
  coord_cartesian(ylim = c(0, 15), clip = "off") +
  theme_classic(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 8),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(face = "bold", size = 9),
    axis.title = element_text(face = "bold", size = 10),
    axis.text.x = element_text(face = "bold", size = 7, angle = 35, hjust = 1),
    axis.text.y = element_text(face = "bold", size = 7),
    axis.line  = element_line(linewidth = 0.6),
    axis.ticks = element_line(linewidth = 0.6),
    legend.title = element_text(face = "bold", size = 9),
    legend.text  = element_text(size = 8)
  )

# print(p_jitter)






```


```{r top_shared_upregulated_VST_jitter_pub_new, message=FALSE, warning=FALSE, fig.width=6}
print(p_jitter)

```
## 9) Gene set enrichment analysis (GSEA): Hallmark and REACTOME

This section runs **pre-ranked GSEA** on the *gene* results (excluding TE features) for both cohorts, using MSigDB **Hallmark** and **C2:CP:REACTOME** gene sets.

```{r gsea-hallmark-reactome, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(ggplot2)
  library(msigdbr)
  library(fgsea)
})

dir.create("outputs/gsea", showWarnings = FALSE, recursive = TRUE)

# ---- sanity checks ----
if (!exists("fit_c1") || !exists("fit_c2")) {
  stop("GSEA chunk needs fit_c1 and fit_c2 (from the Discovery/Validation DESeq2 chunks). ",
       "Please run the Discovery and Validation sections first.")
}

# ---- helper: extract gene-only DESeq2 result table ----
get_gene_res <- function(res) {
  df <- as.data.frame(res) %>% tibble::rownames_to_column("feature")
  df %>%
    filter(!grepl("^TEF_", feature)) %>%          # drop TE rows
    filter(!is.na(stat))                          # fgsea needs numeric ranks
}

# ---- helper: build rank vector (named) ----
# We rank by DESeq2 'stat' (Wald statistic): standard for pre-ranked GSEA
make_ranks <- function(gene_df) {
  ranks <- gene_df$stat
  names(ranks) <- gene_df$feature
  ranks <- ranks[is.finite(ranks)]
  ranks <- sort(ranks, decreasing = TRUE)
  ranks
}

# ---- optional: map Ensembl IDs -> HGNC symbols if needed ----
maybe_map_ensembl_to_symbol <- function(ranks) {
  # heuristic: Ensembl IDs often start with ENSG
  if (!any(grepl("^ENSG", names(ranks)))) return(ranks)

  if (!requireNamespace("AnnotationDbi", quietly = TRUE) ||
      !requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    stop("Your gene IDs look like Ensembl (ENSG...), but packages org.Hs.eg.db and AnnotationDbi ",
         "are not installed. Install them or provide gene symbols in the count matrix rownames.")
  }

  key <- sub("\\..*$", "", names(ranks))  # strip version suffix if present
  map <- AnnotationDbi::select(org.Hs.eg.db::org.Hs.eg.db,
                               keys = unique(key),
                               keytype = "ENSEMBL",
                               columns = c("SYMBOL"))
  map <- map %>%
    filter(!is.na(SYMBOL), SYMBOL != "") %>%
    distinct(ENSEMBL, SYMBOL)

  # aggregate duplicate symbols by max |stat|
  tmp <- tibble(ENSEMBL = key, stat = as.numeric(ranks)) %>%
    left_join(map, by = "ENSEMBL") %>%
    filter(!is.na(SYMBOL)) %>%
    group_by(SYMBOL) %>%
    summarise(stat = stat[which.max(abs(stat))], .groups = "drop")

  out <- tmp$stat
  names(out) <- tmp$SYMBOL
  out <- sort(out, decreasing = TRUE)
  out
}

# ---- MSigDB gene sets ----
msig_h <- msigdbr(species = "Homo sapiens", category = "H") %>%
  select(gs_name, gene_symbol) %>%
  distinct()

msig_react <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME") %>%
  select(gs_name, gene_symbol) %>%
  distinct()

# ---- run fgsea for one cohort ----
run_fgsea <- function(ranks, pathways_df, label, minSize = 15, maxSize = 500) {
  pathways <- split(pathways_df$gene_symbol, pathways_df$gs_name)
  fgsea::fgseaMultilevel(
    pathways = pathways,
    stats = ranks,
    minSize = minSize,
    maxSize = maxSize,
    eps = 0.0
  ) %>%
    as_tibble() %>%
    arrange(padj, desc(abs(NES))) %>%
    mutate(Cohort = label)
}

# ---- build ranks for each cohort ----
gene_c1 <- get_gene_res(fit_c1$res)
gene_c2 <- get_gene_res(fit_c2$res)

ranks_c1 <- make_ranks(gene_c1) %>% maybe_map_ensembl_to_symbol()
ranks_c2 <- make_ranks(gene_c2) %>% maybe_map_ensembl_to_symbol()

# ---- GSEA: Hallmark ----
gsea_h_c1 <- run_fgsea(ranks_c1, msig_h, "Discovery")
gsea_h_c2 <- run_fgsea(ranks_c2, msig_h, "Validation")

gsea_h_all <- bind_rows(gsea_h_c1, gsea_h_c2)

# write.csv(gsea_h_all, "outputs/gsea/GSEA_Hallmark_Discovery_Validation.csv", row.names = FALSE)

# ---- GSEA: REACTOME ----
gsea_r_c1 <- run_fgsea(ranks_c1, msig_react, "Discovery")
gsea_r_c2 <- run_fgsea(ranks_c2, msig_react, "Validation")

gsea_r_all <- bind_rows(gsea_r_c1, gsea_r_c2)

# write.csv(gsea_r_all, "outputs/gsea/GSEA_REACTOME_Discovery_Validation.csv", row.names = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})

# ------------------------------------------------------------
# Dotplot helper (works for fgsea outputs: requires column 'pathway')
# Expects gsea_df has at least: pathway, NES, padj, Cohort
# ------------------------------------------------------------
plot_gsea_dot <- function(gsea_df, title, top_n = 15) {

  req <- c("pathway", "NES", "padj", "Cohort")
  miss <- setdiff(req, colnames(gsea_df))
  if (length(miss) > 0) {
    stop("plot_gsea_dot(): gsea_df is missing columns: ", paste(miss, collapse = ", "))
  }

  df_plot <- gsea_df %>%
    filter(!is.na(padj), is.finite(NES)) %>%
    group_by(Cohort) %>%
    arrange(padj, .by_group = TRUE) %>%
    slice_head(n = top_n) %>%
    ungroup() %>%
    mutate(
      gs_name = pathway,
      gs_name = gsub("^HALLMARK_", "", gs_name),
      gs_name = gsub("^REACTOME_", "", gs_name),
      gs_name = gsub("_", " ", gs_name),
      neglog10padj = -log10(padj)
    )

  ggplot(df_plot, aes(x = NES, y = reorder(gs_name, NES))) +
    geom_point(aes(size = neglog10padj, color = NES), alpha = 0.85) +
    facet_grid(Cohort ~ ., scales = "free_y") +
    scale_color_gradient2(low = "#0072B2", mid = "white", high = "#D55E00", midpoint = 0) +
    labs(
      title = title,
      x = "Normalized Enrichment Score (NES)",
      y = NULL,
      size = "-log10(adj p)",
      color = "NES"
    ) +
    theme_classic(base_size = 12) +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(face = "bold"),
      axis.text  = element_text(face = "bold", size = 10),
      axis.title = element_text(face = "bold")
    )
}

# ------------------------------------------------------------
# Safety checks: you should already have these from the GSEA chunk
# ------------------------------------------------------------
need_objs <- c("gsea_h_all", "gsea_r_all")
missing <- need_objs[!sapply(need_objs, exists)]
if (length(missing) > 0) {
  stop("Missing objects: ", paste(missing, collapse = ", "),
       ".\nRun the GSEA computation chunk first (fgsea results binding).")
}

# ------------------------------------------------------------
# Hallmark dotplot
# ------------------------------------------------------------
p_h <- plot_gsea_dot(gsea_h_all, "GSEA (Hallmark): Discovery vs Validation", top_n = 15)
# print(p_h)

# ------------------------------------------------------------
# REACTOME dotplot
# ------------------------------------------------------------
p_r <- plot_gsea_dot(gsea_r_all, "GSEA (REACTOME): Discovery vs Validation", top_n = 15)
# print(p_r)

# ------------------------------------------------------------
# Save
# ------------------------------------------------------------
dir.create("outputs/gsea", showWarnings = FALSE, recursive = TRUE)
ggsave("outputs/gsea/GSEA_Hallmark_dotplot.png", p_h, width = 8.5, height = 7.5, dpi = 300)
ggsave("outputs/gsea/GSEA_REACTOME_dotplot.png", p_r, width = 8.5, height = 7.5, dpi = 300)

```


```{r gsea_hallmark_dotplots_split, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})

# ------------------------------------------------------------
# Dotplot helper (Hallmark only). Expects fgsea output columns:
# pathway, NES, padj, Cohort  (Cohort = "Discovery" / "Validation")
# ------------------------------------------------------------
plot_gsea_hallmark_dot <- function(gsea_df, cohort_name, top_n = 15) {

  req <- c("pathway","NES","padj","Cohort")
  miss <- setdiff(req, colnames(gsea_df))
  if (length(miss) > 0) {
    stop("plot_gsea_hallmark_dot(): gsea_df is missing columns: ", paste(miss, collapse = ", "))
  }

  df_plot <- gsea_df %>%
    filter(Cohort == cohort_name, !is.na(padj), is.finite(NES)) %>%
    arrange(padj) %>%
    slice_head(n = top_n) %>%
    mutate(
      gs_name = gsub("^HALLMARK_", "", pathway),
      gs_name = gsub("_", " ", gs_name),
      neglog10padj = -log10(padj)
    )

  ggplot(df_plot, aes(x = NES, y = reorder(gs_name, NES))) +
    geom_point(aes(size = neglog10padj, color = NES), alpha = 0.85) +
    scale_color_gradient2(low = "#0072B2", mid = "white", high = "#D55E00", midpoint = 0) +
    labs(
      title = paste0("GSEA (Hallmark): ", cohort_name),
      subtitle = paste0("Top ", top_n, " pathways by adjusted p-value"),
      x = "Normalized Enrichment Score (NES)",
      y = NULL,
      size = "-log10(adj p)",
      color = "NES"
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(face = "bold", size = 10),
      axis.text  = element_text(face = "bold", size = 10),
      axis.title = element_text(face = "bold")
    )
}

# ------------------------------------------------------------
# Safety check: object from your GSEA computation chunk
# ------------------------------------------------------------
if (!exists("gsea_h_all")) {
  stop("Object 'gsea_h_all' not found. Run the Hallmark GSEA computation chunk first.")
}

# Optional: if Cohort column is missing but you have separate objects, stop clearly
if (!("Cohort" %in% colnames(gsea_h_all))) {
  stop("'gsea_h_all' must contain a column named 'Cohort' with values 'Discovery'/'Validation'.")
}

# ------------------------------------------------------------
# Make separate plots
# ------------------------------------------------------------
p_h_disc <- plot_gsea_hallmark_dot(gsea_h_all, cohort_name = "Discovery",  top_n = 15)
p_h_val  <- plot_gsea_hallmark_dot(gsea_h_all, cohort_name = "Validation", top_n = 15)

# print(p_h_disc)
# print(p_h_val)

# ------------------------------------------------------------
# Save
# ------------------------------------------------------------
dir.create("outputs/gsea", showWarnings = FALSE, recursive = TRUE)
ggsave("outputs/gsea/GSEA_Hallmark_dotplot_Discovery.png",  p_h_disc, width = 8.5, height = 7.5, dpi = 300)
ggsave("outputs/gsea/GSEA_Hallmark_dotplot_Validation.png", p_h_val,  width = 8.5, height = 7.5, dpi = 300)
```


```{r gsea_hallmark_combined_deseq2, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(DESeq2)
  library(dplyr)
  library(tibble)
  library(msigdbr)
  library(fgsea)
  library(ggplot2)
})

# ------------------------------------------------------------
# 0) Required inputs
#   - counts_c1, meta_c1 : Discovery
#   - counts_c2, meta_c2 : Validation
#   - meta must contain: response_induction with levels Complete_Response/Refractory
# ------------------------------------------------------------
need_objs <- c("counts_c1","meta_c1","counts_c2","meta_c2")
missing <- need_objs[!sapply(need_objs, exists)]
if (length(missing) > 0) {
  stop("Missing objects: ", paste(missing, collapse = ", "),
       ". Make sure earlier chunks create counts_c1/meta_c1 and counts_c2/meta_c2.")
}

# ------------------------------------------------------------
# Helper: align counts/meta and standardize sample IDs
# ------------------------------------------------------------
align_counts_meta <- function(counts, meta, cohort = "Cohort") {
  counts <- as.matrix(counts)
  meta   <- as.data.frame(meta)

  if (is.null(rownames(meta)) || any(rownames(meta) == "")) {
    if ("Sample" %in% colnames(meta)) rownames(meta) <- meta$Sample
    if ("sample" %in% colnames(meta)) rownames(meta) <- meta$sample
  }

  if (is.null(colnames(counts)) || any(colnames(counts) == "")) {
    stop(cohort, ": counts has missing/empty column names (sample IDs).")
  }
  if (is.null(rownames(meta)) || any(rownames(meta) == "")) {
    stop(cohort, ": meta has missing/empty rownames. Provide rownames(meta)=sample IDs or a Sample column.")
  }

  clean_id <- function(x) gsub("[[:space:]]+", "_", trimws(as.character(x)))
  colnames(counts) <- clean_id(colnames(counts))
  rownames(meta)   <- clean_id(rownames(meta))

  common <- intersect(colnames(counts), rownames(meta))
  if (length(common) == 0) {
    stop(cohort, ": no matching sample IDs between counts colnames and meta rownames.\n",
         "Example counts: ", paste(head(colnames(counts), 5), collapse=", "), "\n",
         "Example meta:   ", paste(head(rownames(meta), 5), collapse=", "))
  }

  counts2 <- counts[, common, drop = FALSE]
  meta2   <- meta[common, , drop = FALSE]
  stopifnot(identical(colnames(counts2), rownames(meta2)))

  storage.mode(counts2) <- "integer"
  list(counts = counts2, meta = meta2)
}

# ------------------------------------------------------------
# 1) Align each cohort, add Cohort column, then combine
# ------------------------------------------------------------
a1 <- align_counts_meta(counts_c1, meta_c1, cohort = "Discovery")
a2 <- align_counts_meta(counts_c2, meta_c2, cohort = "Validation")

meta1 <- a1$meta %>% mutate(Cohort = "Discovery")
meta2 <- a2$meta %>% mutate(Cohort = "Validation")

# Ensure response_induction exists
if (!("response_induction" %in% colnames(meta1)) || !("response_induction" %in% colnames(meta2))) {
  stop("meta_c1/meta_c2 must contain column 'response_induction'.")
}

# Combine counts by columns (samples). Row order must match.
common_rows <- intersect(rownames(a1$counts), rownames(a2$counts))
if (length(common_rows) == 0) stop("No shared features between counts_c1 and counts_c2 rownames.")
counts1 <- a1$counts[common_rows, , drop = FALSE]
counts2 <- a2$counts[common_rows, , drop = FALSE]

counts_all <- cbind(counts1, counts2)
meta_all   <- bind_rows(meta1, meta2)
stopifnot(identical(colnames(counts_all), rownames(meta_all)))

# Standardize factor levels
meta_all <- meta_all %>%
  mutate(
    Cohort = factor(Cohort, levels = c("Discovery","Validation")),
    response_induction = factor(response_induction,
                                levels = c("Complete_Response","Refractory"))
  )

# ------------------------------------------------------------
# 2) DESeq2 on combined cohort (controls for Cohort)
#    Main contrast: Refractory vs Complete_Response
# ------------------------------------------------------------
dds_all <- DESeqDataSetFromMatrix(
  countData = counts_all,
  colData   = meta_all,
  design    = ~ sex + age_dx + response_induction
)

# Optional: remove low-count features for speed/stability
keep <- rowSums(counts(dds_all)) >= 10
dds_all <- dds_all[keep, ]

dds_all <- DESeq(dds_all, quiet = TRUE)

res_all <- results(dds_all, contrast = c("response_induction","Refractory","Complete_Response"))
res_all <- as.data.frame(res_all) %>%
  rownames_to_column("feature") %>%
  mutate(stat = as.numeric(stat))

# ------------------------------------------------------------
# 3) Build ranked list for GSEA (DESeq2 stat)
#    Remove TE features if they are prefixed as TEF_
# ------------------------------------------------------------
rank_df <- res_all %>%
  filter(!is.na(stat), is.finite(stat)) %>%
  filter(!grepl("^TEF_", feature))  # keep genes only

ranks <- rank_df$stat
names(ranks) <- rank_df$feature
ranks <- sort(ranks, decreasing = TRUE)

# ------------------------------------------------------------
# 4) Hallmark pathways (MSigDB) + fgsea
# ------------------------------------------------------------
msig_h <- msigdbr(species = "Homo sapiens", category = "H") %>%
  select(gs_name, gene_symbol)

pathways_h <- split(msig_h$gene_symbol, msig_h$gs_name)

set.seed(1)
gsea_h_combined <- fgsea::fgseaMultilevel(
  pathways = pathways_h,
  stats    = ranks,
  minSize  = 15,
  maxSize  = 500
) %>%
  as_tibble() %>%
  arrange(padj) %>%
  mutate(
    gs_clean = gsub("^HALLMARK_", "", pathway),
    gs_clean = gsub("_", " ", gs_clean),
    neglog10padj = -log10(padj)
  )

# Save table
dir.create("outputs/gsea", showWarnings = FALSE, recursive = TRUE)
# write.csv(gsea_h_combined, "outputs/gsea/GSEA_Hallmark_CombinedCohorts.csv", row.names = FALSE)

# ------------------------------------------------------------
# 5) Simple dotplot for combined GSEA (top 20)
# ------------------------------------------------------------
top_n <- 20
gsea_plot_df <- gsea_h_combined %>%
  filter(!is.na(padj)) %>%
  slice_head(n = top_n)

p_h_combined <- ggplot(gsea_plot_df,
                       aes(x = NES, y = reorder(gs_clean, NES))) +
  geom_point(aes(size = neglog10padj, color = NES), alpha = 0.85) +
  scale_color_gradient2(low = "#0072B2", mid = "white", high = "#D55E00", midpoint = 0) +
  labs(
    title = "GSEA Hallmark Refractory vs CR",
    subtitle = "DESeq2: ~ Cohort + response_induction (Refractory vs Complete_Response)",
    x = "Normalized Enrichment Score (NES)",
    y = NULL,
    size = "-log10(adj p)",
    color = "NES"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(face = "bold", size = 10),
    axis.text  = element_text(face = "bold", size = 10),
    axis.title = element_text(face = "bold")
  )

ggsave("outputs/gsea/GSEA_Hallmark_CombinedCohorts_dotplot.png",
       p_h_combined, width = 8.5, height = 7.5, dpi = 300)
```

```{r gsea_hallmark_combined_deseq2_new, message=FALSE, warning=FALSE, fig.width=6.5 }
print(p_h_combined)

```



### How to run

1. Save this file as `Section2_TE_volcano.Rmd` (already provided for you here)
2. Edit the 5 URLs in **“Auto-download input files”**
3. Knit to HTML (RStudio: **Knit**), or run:

```r
rmarkdown::render("Section2_TE_volcano.Rmd", output_format = "html_document")
```
